Outline paper - Multiple imputation for statistical disclosure control: Creating synthetic datasets with *mice*

### 2021-09-24 mice_synthesizing/paper.Rmd
Gerko's update

Open science, including open data, has been marked as the future of science [@gewin_data_2016], and the advantages of publicly available research data are numerous [@molloy_open_2011; @walport_brest_sharing_2011]. Collecting research data requires an enormous investment both in terms of time and monetary resources. Openly accessible research data bears the potential of increasing the scientific returns for the same data collection effort. Additionally, the fact that public funds are used for data collection results in increasing demand for the collected data. Nevertheless, the possibilities to distribute research data directly are often very limited due to restrictions with regard to privacy and confidentiality. Although these regulations are much needed, privacy constraints are also ranked among the toughest challenges to overcome in the advancement of computational social science [@lazer_life_2009].

Anonimizing the data might seem a quick and appealing approach to limit the identifyability of participants. However, this approach is not sufficient to fulfil contemporary privacy and confidentiality requirements [@ohm_broken_2009; @national_putting_2007]. Over the years, several other techniques have been used to increase the confidentiality of the data, such as categorizing continuous variables, top coding values above a certain threshold or adding random noise to the observed values [@drechsler_synthetic_2011]. However, these methods may distort relationships between variables, thereby reducing the data quality and the scientific returns for re-using the same data for further research.

An alternative solution has been proposed separately by @rubin_statistical_disclosure_1993 and @little_statistical_1993. Although their approaches differed to some extent, the overarching procedure is to use actually observed data to generate multiply imputed synthetic datasets that can be freely disclosed. Based on some imputation model, values from the actually observed data are replaced by multiple draws from the posterior predictive distribution of the observed data. Using this approach, the researcher could replace the observed dataset as a whole with multiple synthetic versions. Alternatively, the researcher could opt to only replace a subset of the observed data. The values that are to be replaced can be complete variables that could be compared with publicly available datasets or registers, such as addresses, or can be certain values that bear a high risk of being disclosive, such as income values above a certain threshold. [**aren't synthetic values then not also too disclosive?**] <!-- I don't think so, because the synthetic values can only be linked to observed values from other sources if they are exactly equal, or if they are the only value within a certain range. The latter might be possible when the number of values is very small, or when the observed value is so extreme that synthetic versions will still be so extreme. However, if the other values are synthetic as well, it is quite likely that synthetic information does not apply to the real-world observation. I think that the highest risk is in the possibility to extract a range of possible values, which might be uniquely match some other record that is available. -->

Conceptually, the synthetic data framework is based upon the building blocks of multiple imputation of missing data, as proposed by @rubin_multiple_1987. Instead of replacing just the missing values with multiple draws from the posterior predictive distribution, one could easily *overimpute* any observed sensitive values. Similarly to multiple imputation of missing data, the multiple synthetic datasets allow for correct statistical inferences, **regardless of the fact that the "true" value is not included in the analyses**. The analyses over multiple synthetic datasets should be pooled into a single inference, so that the researcher can draw valid conclusions from the pooled results. To that respect, the variance should reflect the added variability that is induced by the imputation procedure. 
Potentially, this approach could fulfill the needs for openly accessibly data, without running into barriers with regard to privacy and confidentiality constraints. However, there is no such thing as a free lunch: data collectors have to put effort in creating high-quality synthetic data. Also, the quality of the synthetic data is highly dependent on the imputation models, and using flawed models to generate synthetic data might bias subsequent analyses. Conversely, if the models used to create the synthetic data are able to preserve the relationships between the variables as in the original data, the synthetic data can be nearly as informative as the observed data. Thus, to fully exploit the benefits of synthetic data, the effort to actually create these high-quality datasets should be kept at a minimum.

To mitigate additional effort of creating synthetic datasets on behalf of the researcher, software aimed at multiple imputation of missing data can be employed. Especially if researchers used this software at an earlier stage in the research process, or acquired familiarity with it during earlier projects, the additional burden of creating synthetic datasets is relatively small. The R-package `mice` [@mice] implements multiple imputation of missing data in a straightforward and user-friendly manner. However, the functionality of `mice` is not restricted to the imputation of missing data, but allows to impute any value in the data. Consequently, `mice` can be utilized for the creation of multiply imputed synthetic datasets.

After creating the multiply imputed synthetic datasets, the goal is to obtain valid statistical inferences. In the missing data framework, this is done by performing statistical analyses on all imputed datasets, and pooling the results of the analyses according to Rubin's rules [-@rubin_multiple_1987]. In the synthetic data framework, the same procedure is followed. However, valid statistical results can only be obtained under the premise that the statistical properties of the analyses are correct. Analyzing synthetic data specifically requires appropriate variance estimates. Notably, the variance estimator that is suitable for analyzing multiply imputed datasets for missing data is inappropriate when multiply imputed synthetic datasets are analyzed. As a result, multiple variance estimators have been proposed to make correct inferences from multiply imputed synthetic datasets. These estimators differ according to whether the data consists entirely of synthetic values, or only a subset of the data is synthetic.

The remainder of this paper is organized as follows. First, the `mice` algorithm for the creation of synthetic data will be shortly explained. The aim is to generate synthetic sets that reassure the privacy and confidentiality of the participants. Second, a straightforward workflow for imputation of synthetic data with `mice` will be demonstrated. The third step is to summarize the variance estimators that have been proposed in the past. Finally, it will be assessed which variance estimator provides valid inferences from the synthetic data.

@@

The `mice` package has been developed for multiple imputation of missing data. In this context, the aim is to replace missing values due to nonresponse by plausible values from the posterior predictive distribution of the variable containing the missings. Doing so, `mice` makes use of fully conditional specification [FCS; @vanbuuren_fully_2006], which breaks down the multivariate distribution of the data $\textbf{Y} = (\textbf{Y}_{obs}, \textbf{Y}_{mis})$ into $j = 1, 2, \dots, k$ univariate conditional densities, with $k$ the number of variables in the data. Using FCS, for every variable containing missing values a model is constructed, to impute the missing values $Y_{j, mis}$ with draws from the posterior predictive distribution of $P(Y_{j, mis} | \textbf{Y}_{obs}, \theta)$ on a variable by variable basis. Note that the predictor matrix $Y_{-j}$ may contain yet imputed values from an earlier imputation step, and thus will be updated after every iteration. This procedure is applied $m$ times, resulting in $m$ completed datasets $\textbf{D} = (\textbf{D}^{(1)}, \textbf{D}^{(2)}, \dots, \textbf{D}^{(m)})$, with $\textbf{D}^{(l)} = (\textbf{Y}_{obs}, Y^{(l)}_{mis})$.

This approach can be straightforwardly extended to a synthetic data framework. However, rather than imputing missing data, the actually observed data is overimputed, that is, the observed values are replaced by multiple imputations. For simplicity, assume that the data is completely observed (i.e., $\textbf{Y} = \textbf{Y}_{obs}$). Following the notation of @reiter_raghunathan_multiple_2007, let $Z_i = 1$ if any of the values of unit $i$ are to be replaced by imputations, and $Z_i = 0$ otherwise, with $\textbf{Z} = (Z_1, Z_2, \dots, Z_n)$. Accordingly, the data consists of values that are to be replaced and values that are to be kept (i.e., $\textbf{Y} = (\textbf{Y}_{rep}, \textbf{Y}_{nrep})$. Now, instead of imputing $\textbf{Y}_{mis}$ with draws from the posterior predictive distribution of $P(Y_{j, mis} | \textbf{Y}_{obs}, \theta)$ as in the missing data case, $\textbf{Y}_{rep}$ is imputed from the posterior distribution of $P(Y^{(l)}_{j, rep} | \textbf{Y}^{(l)}_{-j}, \textbf{Z}, \theta)$, where $l$ is an indicator for the synthetic dataset ($l = 1, 2, \dots, m$). Note that synthetic values that are imputed at an earlier step can be used for imputing variable $j$. This process results in the synthetic data $\textbf{D} = (\textbf{D}^{(1)}, \textbf{D}^{(2)}, \dots, \textbf{D}^{(m)})$.

By default, `mice` imputes all values that are coded as missing in the dataset. However, using the `where` parameter within the `mice`-function, one can choose to impute any value in the dataset. As input, the `where` parameter requires a matrix of the same dimensions as the data, (i.e., a $n \times k$ matrix, with $n$ the number of observations in the collected data) containing logicals $z_{ij}$ that indicate which cells are selected to have their values replaced. This approach allows to *overimpute* a subset of the observed data, or the observed data as a whole, resulting in a dataset that partly or completely consists of synthetic values. Other parameters that are required to be specified are the imputation method (`method`), the number of imputations ($m$) and the predictor matrix, specifying the predictor variables that are used in the imputation model for every variable $j$ that has some of its values replaced (`predictorMatrix`).

Choosing an adequate method to impute the data is paramount, as a flawed imputation model will drastically reduce the data utility in further inferences. Imputation models should be as flexible as possible to capture most of the patterns in the data, and to model, possibly unanticipated, data characteristics [@murray_multiple_2018; @rubin_18years_1996]. Parametric methods, albeit easy to implement in practice, may be too restrictive to capture generally complex patterns in the data, especially in the case of nonlinear relations and interactions between multiple variables. Classification and regression trees [CART; @breiman_cart_1984] allow to model more complex patterns in the data, and have therefore been suggested as an appropriate imputation method [@reiter_cart_2005; @burgette_reiter_cart_2010; @doove_buuren_recursive_2014]. Loosely speaking, CART sequentially splits the predictor space into non-overlapping regions in such a way that the within-region variance is as small as possible after every split. As such, CART does not impose any parametric distribution on the data, making it a widely applicable method that allows for a large variety of relationships within the data [@islr_2013]. Given these appealing characteristics and the call for the use of flexible methods when multiply imputing data, the performance of `mice` with CART as the specified imputation method will be illustrated.

@@

Additional parameters that come with the use of CART are the complexity parameter `cp` and the minimum number of observations in any terminal node `minbucket`, that both restrain the flexibility of the imputation model. Appropriate values for these parameters, as well as the input for the `predictorMatrix`, depend on the data at hand. Hence, the specification of these parameters will be illustrated using an actual dataset. The `boys` data [@fredriks_boys_2000] that is in-built within the `mice` package is used to illustrate the specification of these parameters accordingly. **DETERMINE WHETHER THIS DATASET SUFFICES; BECAUSE IMPUTATION MIGHT MAKE THE DATA ARTIFICIALLY SIMILAR BEFORE SYNTHESIZING, MAKING SYNTHESIZING TOO EASY** Unfortunately, this dataset does not differ from the vast majority of collected datasets, in the sense that it suffers from missing data. For simplicity, the data is completed using the default `mice` imputation model to create a single, fully observed dataset. The data consists of the values of $`r nrow(mice::boys)`$ Dutch boys on the following $`r ncol(mice::boys)`$ variables: age (in years), height (in cm), weight (in kg), body mass index (BMI), head circumference (in cm), Genital Tanner stage (G1-G5), testicular volume (in ml) and region.

>>>>

Open science, including open data, has been marked as the future of science [@gewin_data_2016], and the advantages of publicly available research data are numerous [@molloy_open_2011; @walport_brest_sharing_2011]. Collecting research data requires an enormous investment both in terms of time and monetary resources. Openly accessible research data bears the potential of increasing the scientific returns for the same data collection effort. Additionally, the fact that public funds are used for data collection results in increasing demand for the collected data. Nevertheless, the possibilities to distribute research data directly are often very limited due to restrictions on data privacy and data confidentiality. Although these regulations are much needed, privacy constraints are also ranked among the toughest challenges to overcome in the advancement of computational social science [@lazer_life_2009].

Anonymizing the data might seem a quick and appealing approach to limit the unique identification of participants. However, this approach is not sufficient to fulfill contemporary privacy and confidentiality requirements [@ohm_broken_2009; @national_putting_2007]. Over the years, several other techniques have been used to increase the confidentiality of the data, such as categorizing continuous variables, top coding values above an upper bound or adding random noise to the observed values [@drechsler_synthetic_2011]. However, these methods may distort the true data relation between variables, thereby reducing the data quality and the scientific returns for re-using the same data for further research.

An alternative solution has been proposed separately by @rubin_statistical_disclosure_1993 and @little_statistical_1993. Although their approaches differ to some extent, the overarching procedure is to use bonafide observed data to generate multiply imputed synthetic data sets that can be freely disclosed. Observed data values are then replaced by multiple draws from the posterior predictive distribution of the observed data, based on some imputation model. Using this approach, the researcher could replace the observed data set as a whole with multiple synthetic versions. Alternatively, the researcher could opt to only replace a subset of the observed data. For example, one can choose to only replace dimensions in the data that could be compared with publicly available data sets or registers. Likewise, synthetisation could be limited to those values that are disclosive, such as high incomes or high turnovers. <!-- I don't think so, because the synthetic values can only be linked to observed values from other sources if they are exactly equal, or if they are the only value within a certain range. The latter might be possible when the number of values is very small, or when the observed value is so extreme that synthetic versions will still be so extreme. However, if the other values are synthetic as well, it is quite likely that synthetic information does not apply to the real-world observation. I think that the highest risk is in the possibility to extract a range of possible values, which might be uniquely match some other record that is available. -->

Conceptually, the synthetic data framework is based upon the building blocks of multiple imputation of missing data, as proposed by @rubin_multiple_1987. Instead of replacing just the missing values with multiple draws from the posterior predictive distribution, one could easily *overimpute* any observed sensitive values. Similarly to multiple imputation of missing data, the multiple synthetic datasets allow for correct statistical inferences, **regardless of the fact that the "true" value is not included in the analyses**. The analyses over multiple synthetic data sets should be pooled into a single inference, so that the researcher can draw valid conclusions from the pooled results. To that respect, the variance should reflect the added variability that is induced by the imputation procedure. 
Potentially, this approach could fulfill the needs for openly accessibly data, without running into barriers with regard to privacy and confidentiality constraints. However, there is no such thing as a free lunch: data collectors have to put effort in creating high-quality synthetic data. Also, the quality of the synthetic data is highly dependent on the imputation models, and using flawed models to generate synthetic data might bias subsequent analyses. Conversely, if the models used to create the synthetic data are able to preserve the relationships between the variables as in the original data, the synthetic data can be nearly as informative as the observed data. Thus, to fully exploit the benefits of synthetic data, the effort to actually create these high-quality data sets should be kept at a minimum.

To mitigate additional effort of creating synthetic data sets on behalf of the researcher, software aimed at multiple imputation of missing data can be employed. Especially if researchers used this software at an earlier stage in the research process, or acquired familiarity with it during earlier projects, the additional burden of creating synthetic data sets is relatively small. The R-package `mice` [@mice] implements multiple imputation of missing data in a straightforward and user-friendly manner. However, the functionality of `mice` is not restricted to the imputation of missing data, but allows to impute any value in the data: even observed values. Consequently, `mice` can be utilized for the creation of multiply imputed synthetic data sets.

After creating the multiply imputed synthetic data sets, the goal is to obtain valid statistical inferences in the spirit of @rubin_multiple_1987 and @neyman1934. In the missing data framework, this is done by performing statistical analyses on all imputed datasets, and pooling the results of the analyses according to Rubin's rules [@rubin_multiple_1987, pp.76]. In the synthetic data framework, the same procedure is followed, but with a slight twist: there are no values that remain constant over the synthetic data sets. The procedure of drawing valid inferences from multiple synthetic data sets is therefore slightly different.

In this manuscript we detail a workflow for synthesizing data with `mice`. 
First, the `mice` algorithm for the creation of synthetic data will be shortly explained. The aim is to generate synthetic sets that reassure the privacy and confidentiality of the participants. Second, a straightforward workflow for imputation of synthetic data with `mice` will be demonstrated. Third, we demonstrate the validity of the procedure through statistical simulation.

@@

The `mice` package [@mice] in `R` [@Rproject] has been developed for multiple imputation of missing data. In this context, the aim is to replace missing values due to nonresponse by plausible values from the posterior predictive distribution of the variable containing the missings. Doing so, `mice` makes use of fully conditional specification [FCS; @vanbuuren_fully_2006], which breaks down the multivariate distribution of the data $\textbf{Y} = (\textbf{Y}_{obs}, \textbf{Y}_{mis})$ into $j = 1, 2, \dots, k$ univariate conditional densities, where $k$ denotes the number of columns in the data. Using FCS, a model is constructed for every incomplete variable and the missing values $Y_{j, mis}$ are then imputed with draws from the posterior predictive distribution of $P(Y_{j, mis} | \textbf{Y}_{obs}, \theta)$ on a variable-by-variable basis. Note that the predictor matrix $Y_{-j}$ may contain yet imputed values from an earlier imputation step, and thus will be updated after every iteration. This procedure is applied $m$ times, resulting in $m$ completed datasets $\textbf{D} = (\textbf{D}^{(1)}, \textbf{D}^{(2)}, \dots, \textbf{D}^{(m)})$, with $\textbf{D}^{(l)} = (\textbf{Y}_{obs}, Y^{(l)}_{mis})$.

In `mice`, the generation of multiply imputed data sets is straightforward. The following pseudocode details the multiple imputation of the `mice::boys` data set [@fredriks_boys_2000] into the object `imp` with `m = 10` imputated sets and `maxit = 7` iterations for the algorithm to converge, using the default imputations methods for each column data class.

```{r eval = FALSE}
library(mice)
imp <- mice(boys, 
            m = 10,
            maxit = 7)
```

It is straightforward to extended the imputation approach to generate synthetic values. Rather than imputing missing data, the observed values are then replaced by synthetic draws from the posterior predictive distribution. For simplicity, assume that the data is completely observed (i.e., $\textbf{Y} = \textbf{Y}_{obs}$). Following the notation of @reiter_raghunathan_multiple_2007, let for $n$ units denote $Z_i = 1$ if any of the values of unit $i = 1, 2, \dots, n$ are to be replaced by imputations, and $Z_i = 0$ otherwise, with $\textbf{Z} = (Z_1, Z_2, \dots, Z_n)$. Accordingly, the data consists of values that are to be replaced and values that are to be kept (i.e., $\textbf{Y} = (\textbf{Y}_{rep}, \textbf{Y}_{nrep})$. Now, instead of imputing $\textbf{Y}_{mis}$ with draws from the posterior predictive distribution of $P(Y_{j, mis} | \textbf{Y}_{obs}, \theta)$ as in the missing data case, $\textbf{Y}_{rep}$ is imputed from the posterior distribution of $P(Y^{(l)}_{j, rep} | \textbf{Y}^{(l)}_{-j}, \textbf{Z}, \theta)$, where $l$ is an indicator for the synthetic dataset ($l = 1, 2, \dots, m$). Note that synthetic values that are imputed at an earlier step can be used for imputing variable $j$. This process results in the synthetic data $\textbf{D} = (\textbf{D}^{(1)}, \textbf{D}^{(2)}, \dots, \textbf{D}^{(m)})$.

For example, overimputing the observed and missing values in the `mice::boys` data set into the object `syn`, given the same imputation parameters as before, can be realized by the following code execution

```{r eval = FALSE}
syn <- mice(boys, 
            m = 10,
            maxit = 7, 
            where = matrix(TRUE, 
                           nrow = nrow(boys),
                           ncol = ncol(boys)))
```

where the argument `where` requires a matrix of the same dimensions as the data, (i.e., a $n \times k$ matrix) containing logicals $z_{ij}$ that indicate which cells are selected to have their values replaced by draws from the posterior predictive distribution. This approach allows to *overimpute* a subset of the observed data, or - as in the above example - the observed data as a whole, resulting in a dataset that partly or completely consists of synthetic data values.

Choosing an adequate imputation model to impute the data is paramount, as a flawed imputation model may drastically impact the validity of inferences. Imputation models should be as flexible as possible to capture most of the patterns in the data, and to model possibly unanticipated data characteristics [@murray_multiple_2018; @rubin_18years_1996]. Parametric methods, albeit easy to implement in practice, may be too restrictive to capture generally complex patterns in the data, especially in the case of nonlinear relations and interactions between multiple variables. Classification and regression trees [CART; @breiman_cart_1984] allow to model more complex patterns in the data, and have therefore been suggested as an appropriate imputation method [@reiter_cart_2005; @burgette_reiter_cart_2010; @doove_buuren_recursive_2014]. Loosely speaking, CART sequentially splits the predictor space into non-overlapping regions in such a way that the within-region variance is as small as possible after every split. As such, CART does not impose any parametric distribution on the data, making it a widely applicable method that allows for a large variety of relationships within the data [@islr_2013]. Given these appealing characteristics and the call for the use of flexible methods when multiply imputing data, we will focus our illustrations and evaluations of `mice` to method `mice.impute.cart()`, realized by

```{r eval = FALSE}
syn <- mice(boys, 
            m = 10,
            maxit = 7, 
            method = "cart",
            where = matrix(TRUE, 
                           nrow = nrow(boys),
                           ncol = ncol(boys)))
```

@@

Additional parameters that come with the use of `mice.impute.cart()` are the complexity parameter `cp` and the minimum number of observations in any terminal node `minbucket`, that both restrain the flexibility of the imputation model. Appropriate values for these parameters, as well as the input for the `predictorMatrix`, depend on the data at hand. Hence, the specification of these parameters will be illustrated using an actual dataset. The `boys` data [@fredriks_boys_2000] that is in-built within the `mice` package is used to illustrate the specification of these parameters accordingly. **DETERMINE WHETHER THIS DATASET SUFFICES; BECAUSE IMPUTATION MIGHT MAKE THE DATA ARTIFICIALLY SIMILAR BEFORE SYNTHESIZING, MAKING SYNTHESIZING TOO EASY** Unfortunately, this dataset does not differ from the vast majority of collected datasets, in the sense that it suffers from missing data. For simplicity, the data is completed using the default `mice` imputation model to create a single, fully observed dataset. The data consists of the values of $`r nrow(mice::boys)`$ Dutch boys on the following $`r ncol(mice::boys)`$ variables: age (in years), height (in cm), weight (in kg), body mass index (BMI), head circumference (in cm), Genital Tanner stage (G1-G5), testicular volume (in ml) and region.
